Human Factors Influencing Visual Statistical Inference

### 2014-10-21 sociology_paper.tex
edits throughout section 1

\hh{How do we interpret this finding, though?} For that, we need to know the context of the data and we need to have more information about \hh{how the null plots were generated}. 
This particular example investigates the results from the 2012 US presidential election in comparison to  poll results just prior to the election. 
Although this example is more simplistic than most of the tests conducted to date, it will serve the purpose of illustrating the lineup protocol. The data consists of the difference in poll results between the two (major) presidential candidates, Obama and Romney, for all states. Each panel in Figure~\ref{fig:elect-1} shows an `electoral building' \citep{mosley2010} where each state in the union is represented by a rectangle. This difference is plotted horizontally, and the height of each box corresponds to the state's electoral votes. Color indicates party affiliation.  The null hypothesis is ``that the election results were consistent with the polls''. The polling results provides the null model from which data is simulated. \hh{A normal model with mean and standard deviation based on a poll and its margin of error is used to} simulate different scenarios that might have resulted on election day, if the polls were on target. A null data set is generated as a set of draws from this model. % normal distribution, with mean equal to the difference in poll percentage of the latest state poll results, and standard deviation equal to 2.5, approximating a margin of error of 5\%. 
These samples are plotted as electoral buildings, and the plot made with the results from the election is placed randomly among them in a lineup of size $5\times 4$. If the null hypothesis is true, the actual data plot should look  like any of the other plots, and not be identifiable by an observer. Figure~\ref{fig:tower} shows the plot of the electoral building with added context information and labels.
 %A party needs 270 electoral votes or more to win the presidential election. 
%\begin{figure}[htbp] 
%   \centering
%   \includegraphics[width=0.8\linewidth]{electoral-5-13.pdf} 
%   \caption{Which plot looks the most different from the other plots?  Lineup \#1 - electoral building; another set of four lineups can be found in figs.~{fig:elect.2} through~{fig:elect.5}}
%   \label{fig:elect.1}
%\end{figure}

@@

A lineup can be evaluated by a single person or multiple observers. A distribution \hh{similar to the binomial distribution, but adjusted for dependencies introduced by the lineup scenario,} is used to calculate the $p$-value based on the number of times observers identify the actual data plot, which provides the information needed to make a decision on rejecting or failing to reject the null hypothesis.  In order to avoid expectation errors \citep{meilgaard}, observers should not be aware of the data that constitutes a lineup, and should not have seen the actual data plot before seeing the lineup. This is the reason that in the election example, above, the scenario was explained after the lineup question, in the text.

The question that is asked of the observer should be as general as possible, effectively asking the observer to pick the plot that is different, and allowing them to provide their reasons for seeing their pick as different. In some of the studies, the ones described in  \cite{majumder:2013} very specific questions were asked, because the experiments were being conducted to compare results from the lineup protocol with those of conventional tests. In those experiments, structure in the data was strictly controlled in the simulation process, which allowed for specific questions to be asked. In the election example, observers were asked ``which plot is the most different?''. The type of plot, showing two (modified) stacked bar charts in different colors should suggest to the observer that the interesting feature is most difference between the two heights. Most observers got this, but it is possible that some observers might pick plot 4, where the red tower is slightly above the blue as the most different because it is the only plot with this feature. So a better question may have been ``Which plot shows the biggest height difference between the two towers?'' except that this tailors the inference to a specific feature which does not match the null hypothesis of interest. 
%=======
%A lineup can be evaluated by a single person or multiple observers. Based on the feedback of $K$ observers a $p$-value is computed with the help of a binomial random variable $X$ for the number of observers picking the data plot. Under the null hypothesis is $X$ is  distributed according to $B_{K, 1/20}$ \citep{majumder:2013}. The $p$-value for a lineup is then given as $P(X \ge x)$. 
%%This finally a decision is made about the hypothesis being tested.
%>>>>>>> example

>>>>

\hh{How do we interpret this finding, though?} For that, we need to know the context of the data and we need to have more information about \hh{how the null plots were generated}. 
This particular example investigates the results from the 2012 US presidential election in comparison to  poll results just prior to the election. 
Although this example is more simplistic than most of the tests conducted to date, it will serve the purpose of illustrating the lineup protocol. The data consists of the difference in poll results between the two (major) presidential candidates, Obama and Romney, for all states. Each panel in Figure~\ref{fig:elect-1} shows an `electoral building' \citep{mosley2010} where each state in the union is represented by a rectangle. This difference is plotted horizontally, and the height of each box corresponds to the state's electoral votes. Color indicates party affiliation.  The null hypothesis is ``that the election results were consistent with the polls''. The polling results provides the null model from which data is simulated. \hh{A normal model with mean and standard deviation based on a poll and its margin of error is used to} simulate different scenarios that might have resulted on election day, if the polls were on target. A null data set is generated as a set of draws from this model. % normal distribution, with mean equal to the difference in poll percentage of the latest state poll results, and standard deviation equal to 2.5, approximating a margin of error of 5\%. 
These samples are plotted as electoral buildings, and the plot based on the actual election results is placed randomly among the null plots in a lineup of size $5\times 4$. If the null hypothesis is true, the  plot of the actual data should look  like any of the other plots, and not be identifiable by an observer. Figure~\ref{fig:tower} shows the plot of the electoral building with added context information and labels.
 %A party needs 270 electoral votes or more to win the presidential election. 
%\begin{figure}[htbp] 
%   \centering
%   \includegraphics[width=0.8\linewidth]{electoral-5-13.pdf} 
%   \caption{Which plot looks the most different from the other plots?  Lineup \#1 - electoral building; another set of four lineups can be found in figs.~{fig:elect.2} through~{fig:elect.5}}
%   \label{fig:elect.1}
%\end{figure}

@@

A lineup can be evaluated by a single person or multiple observers. A distribution \hh{similar to the binomial distribution, but adjusted for dependencies introduced by the lineup scenario,} is used to calculate the $p$-value based on the number of times observers identify the actual data plot, which provides the information needed to make a decision on rejecting or failing to reject the null hypothesis.  In order to avoid expectation errors \citep{meilgaard}, observers should not be aware of the data that constitutes a lineup, and should not have seen the actual data plot before seeing the lineup. This is the reason that in \hh{our first example we explained the context only after presenting the lineup.}%the election example, above, the scenario was explained after the lineup question, in the text.

\hh{When presenting a lineup to an observer we should also phrase the accompanying question in as general a manner as possible, effectively}
%The question that is asked of the observer should be as general as possible, 
effectively asking the observer to pick the plot that is \hh{most} different, and allowing \hh{observers} to provide their reasons for their choice. 
\hh{If lineup tests are run in head-to-head comparisons with conventional statistical tests, such as the experiments  in  \cite{majumder:2013}, these questions have to be phrased much more specifically, in order to make the situations more comparable. } 
%In some of the studies, the ones described in  \cite{majumder:2013} very specific questions were asked, because the experiments were being conducted to compare results from the lineup protocol with those of conventional tests. 
In those experiments, the structure in the data is strictly controlled in the simulation process, which allows for specific questions to be asked. 
In the election example, observers were asked ``which plot is the most different?''. The type of plot, showing two (modified) stacked bar charts in different colors should suggest to the observer that the interesting feature is the difference between the two heights. 
\hh{Most observers based their decision on this -- besides  `asymmetry' between the towers (a reason suggested by us), most free-form responses for the reason of choice describe a comparison between the red and blue towers. 
In the responses for this lineup we see something that is very typical of responses in lineups: it is only a few panels from the lineup that observers pick, while most panels do not get picked or get picked only a few times. Out of a total of 72~responses, 23 observers picked panel~\#8, 12 observers picked the (data) panel \#13, 10 observers picked panel~\#4, 6 observers picked panel \#7, and all of the remaining panels were picked at most three times. From the panels picked, we see that the top three choices  are all extreme with respect to heights of blue versus red towers:  both panels \#8 and \#13 show a large difference between the height of the towers, while panel \#4 is the only panel, in which the red tower is higher than the blue one. }
% results from the turk study:
% panel:   1 10 11 12 13 14 15 16 18 19  2 20  3   4  5  6  7  8  9 
% #picks:  2  1    0  1 12    3  3    1  1   1  2   1  0 10  2  2  6 23  1

%Most observers got this, but it is possible that some observers might pick plot~4, where the red tower is slightly above the blue as the most different because it is the only plot with this feature \hh{(10 out of 72 observers actually did this)}. 
  %a better question may

 \hh{If we consider this difference to be the main objective of the visualization, a more focused question would}
 have been ``Which plot shows the biggest height difference between the two towers?'' %except that this tailors the inference to a specific feature which does not match the null hypothesis of interest.

%=======
%A lineup can be evaluated by a single person or multiple observers. Based on the feedback of $K$ observers a $p$-value is computed with the help of a binomial random variable $X$ for the number of observers picking the data plot. Under the null hypothesis is $X$ is  distributed according to $B_{K, 1/20}$ \citep{majumder:2013}. The $p$-value for a lineup is then given as $P(X \ge x)$. 
%%This finally a decision is made about the hypothesis being tested.
%>>>>>>> example
